# (PART) Appendix {-}

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 100)
```

# Reading and writing spatial data in R{-}

## Sample files for this exercise{-}

First, you will need to download some sample files from the github repository.  Make sure to set your R session folder to the directory where you will want to save the sample files before running the following code chunks.


```{r results='hide'}
download.file("http://github.com/mgimond/Spatial/raw/master/Data/Income_schooling.zip", 
              destfile = "Income_schooling.zip" , mode='wb')
unzip("Income_schooling.zip", exdir = ".")
file.remove("Income_schooling.zip")

```


```{r}
download.file("http://github.com/mgimond/Spatial/raw/master/Data/rail_inters.gpkg", 
              destfile = "./rail_inters.gpkg", mode='wb')
```

```{r}
download.file("http://github.com/mgimond/Spatial/raw/master/Data/elev.img",  
              destfile = "./elev.img", mode='wb')               
```

## Introduction {-}

There are several different R spatial formats to choose from. Your choice of format will largely be dictated by the package(s) and or function(s) used in your workflow. A breakdown of formats and intended use are listed below.


```{r echo=FALSE}
my_tbl <- tibble::tribble(
  ~`Data format`, ~`Used with...`, ~`Used in package...`,  ~`Used for...`, ~Comment,
  "\`sf\`", "vector",  "\`sf\`, others", "visualizing, manipulating, querying", "This is likely to become the new spatial standard in R. Will also read from spatially enabled databases such as postgresSQL.",
   "\`raster\`",  "raster", "\`raster\`, others", "visualizing, manipulating, spatial statistics", "This is the most versatile raster format",
   "\`SpatialPoints*\` \n \`SpatialPolygons*\` \n \`SpatialLines*\` \n \`SpatialGrid*\`\n",  "vector and raster",  "\`sp\`, \`spdep\`",  "Visualizing, spatial statistics", "May be superseded by `sf` in the future",
    "\`ppp\` \`owin\`",   "vector",   "\`spatstat\`", "Point pattern analysis/statistics", NA,
     "\`im\`",   "raster",   "\`spatstat\`", "Point pattern analysis/statistics", NA
  )

library(knitr)
library(kableExtra)
library(dplyr)
kable_styling(
        kable(my_tbl, digits = 3, row.names = FALSE, align = c("r", "c","c","l","l") ,
        caption = NULL, format = "html"),
        bootstrap_options = c("striped", "hover", "condensed"),
        position = "left", full_width = FALSE) %>% 
  row_spec(0, color="white", background = "#6E6E6E", align="c") %>% 
  footnote(number=c("The `spatial*` format includes SpatialPointsDataFrame, SpatialPolygonsDataFrame, SpatialLinesDataFrame, etc...") )
```


There is an attempt at standardizing the spatial format in the R ecosystem by adopting a well established set of spatial standards known as [simple features](https://en.wikipedia.org/wiki/Simple_Features). This effort results in a recently developed package called   [`sf`](https://r-spatial.github.io/sf/).

It is therefore recommended that you work in an `sf` framework  when possible. As of this writing, most of the _basic_ data manipulation and visualization operations can be successfully conducted using `sf` spatial objects. 

Some packages such as `spdep` and `spatstat` require specialized data object types. This tutorial will highlight some useful conversion functions for this purpose.



## Creating spatial objects {-}

The following sections demonstrate different spatial data object creation strategies.

### Reading a shapefile {-}

Shapefiles consist of many files sharing the same core filename and different suffixes (i.e. file extensions). For example, the sample shapefile used in this exercise consists of the following files:

```{r echo= FALSE}
list.files(".", "Income_schooling.*")
```

Note that the number of files associated with a shapefile can vary. `sf` only needs to be given the `*.shp` name. It will then know which other files to read into R such as projection information and attribute table.

```{r results='hide'}
library(sf)
s.sf <- st_read("Income_schooling.shp")
```

Let's view the first few records in the spatial data object. 

```{r}
head(s.sf, n=4)  # List spatial object and the first 4 attribute records
```

Note that the `sf` object stores not only the geometry but the coordinate system information and attribute data as well. These will be explored later in this exercise.

### Reading a GeoPackage {-}

A geopackage can store more than one layer. To list the layers available in the geopackage, type:

```{r}
st_layers("rail_inters.gpkg")
```

In this example, we have two separate layers: `Interstate` and `Rail`. We can extract each layer separately via the `layer=` parameter.

```{r results='hide'}
inter.sf <- st_read("rail_inters.gpkg", layer="Interstate")
rail.sf  <- st_read("rail_inters.gpkg", layer="Rail")
```

### Reading a raster {-}

The `raster` package will read many different raster file formats such as geoTiff, Imagine and HDF5 just to name a few. To see a list of supported raster file formats simply run `rgdal::gdalDrivers()` at a command prompt. The `rgdal` package is normally installed with your installation of `raster`.

In the following example, an _Imagine_ raster file is read into R.

```{r}
library(raster)
elev.r <- raster("elev.img")
```

What sets a `raster` object apart from other R data file objects is its storage. By default, data files are loaded into memory but `raster` objects are not. This can be convenient when working with raster files too large to load into memory. But this comes at a performance cost. If your RAM is large enough to handle your raster file, it's best to load the entire dataset into memory.

To check if the `elev.r` object is loaded into memory, run:

```{r}
inMemory(elev.r)
```

To force the raster into memory use `readAll()`:

```{r}
elev.r <- readAll( raster("elev.img") )
```

Let's check that the raster is indeed loaded into memory:

```{r}
inMemory(elev.r)
```

Now let's look at the raster's properties:

```{r}
elev.r
```

The raster object returns its grid dimensions (number of rows and columns), pixel size/resolution (in the layer's coordinate system units), geographic extent, native coordinate system (UTM NAD83 Zone 19 with units of meters) and min/max raster values.

### Creating a spatial object from a data frame {-}

Geographic point data locations recorded in a spreadsheet can be converted to a spatial point object. Note that it's important that you specify the coordinate system used to record the coordinate pairs since such information is not stored in a data frame. In the following example, the coordinate values are recorded in a WGS 1984 geographic coordinate system (`crs = 4326`).

```{r}
# Create a simple dataframe with lat/long values
df <- data.frame(lon = c(-68.783, -69.6458, -69.7653),
                 lat = c(44.8109, 44.5521, 44.3235),
                 Name= c("Bangor", "Waterville", "Augusta"))

# Convert the dataframe to a spatial object. Note that the
# crs= 4326 parameter assigns a WGS84 coordinate system to the 
# spatial object
p.sf <- st_as_sf(df, coords = c("lon", "lat"), crs = 4326) 
p.sf  
```

### Geocoding street addresses {-}

The `ggmap` package offers a geocoding function called `mutate_geocode` which will take a table with physical addresses and create a new table with latitude and longitude values for those addresses. The new table also contains all other columns of the input table. The function makes use of the [*Data Science Toolkit*](http://www.datasciencetoolkit.org/) API.

Let's first create a new data table with addresses for a few liberal arts colleges in the State of Maine.

```{r}
addr <- data.frame(Address = c("4000 mayflower Hill, Waterville, Maine", 
                               "2 Andrews Rd, Lewiston, ME 04240",
                               "255 Maine St, Brunswick, ME 04011",
                               "90 Quaker Hill Rd, Unity, ME 04988",
                               "105 Eden St, Bar Harbor, ME 04609"),
                   College = c("Colby", "Bates", "Bowdoin", "Unity", "CoA"),
                   stringsAsFactors = FALSE)
```

Next, we'll geocode the addresses and save the output to a new data table called `addr.geo`.

```{r}
library(ggmap)
addr.geo <- mutate_geocode(addr, location=Address, output="latlona" , source="dsk")
```

The `location` parameter is assigned the column with the addresses. The `output` parameter is given instructions on _what_ to output. Here, the option `latlona` instructs the function to output both the lat/lon coordinate values as well as all other attributes. It also adds another column called `address` (note the lower case `a`) with the nearest address match. Note that not all addresses are necessarily matched.

```{r}
addr.geo
```

Note that `addr.geo` is not a spatial object; it's a data frame. It can be converted to an `sf` object via `st_as_sf` as outlined in the previous section. The lat/lon values are assumed to be recorded in a WGS84 geographic coordinate system.

```{r}
la.sf <- st_as_sf(addr.geo, coords = c("lon", "lat"), crs = 4326) 
la.sf
```


You can also use the US Census Bureau's [geocoding service](https://geocoding.geo.census.gov/geocoder/locations/addressbatch?form) for creating lat/lon values from US street addresses. This needs to be completed via their website and the resulting data table (a CSV file) would then need to be loaded in R as a data frame.

## Converting from an `sf` object {-}

Packages such as `spdep` and `spatsat` currently do not support `sf` objects. The following sections demonstrate methods to convert from `sf` to other formats.

### Converting an `sf` object to a `Spatial*` object (`spdep`/`sp`) {-}

The following code will convert point, polyline or polygon features to a `spatial*` object. In this example, an `sf` polygon feature is converted to a `SpatialPolygonsDataFrame` object.

```{r}
s.sp <- as(s.sf, "Spatial")
class(s.sp)
```

Note that if you wish to create a `Spatial*` object directly from a shapefile (and bypass the `sf` object creation), you could run the `maptools` function `readShapeSpatial("Income_schooling.shp")`. However, this approach _strips_ the coordinate system information from the spatial object.

### Converting an `sf` polygon object to an `owin` object {-}

The `spatstat` package is normally used to analyze point patterns however, in most cases, the study extent needs to be explicitly defined by a polygon object. The polygon should be of class `owin`. Conversion from `sf` to `owin` requires the use of the `maptools` package.

Note that the attribute table gets stripped from the polygon data. This is usually fine given that the only reason for converting a polygon to an `owin` format is for delineating the study boundary.

```{r}
library(maptools)
s.owin <- as(s.sp, "owin")
class(s.owin)
```

### Converting an `sf` point object to a `ppp` object {-}

As of this writing, it seems that you need to first convert the `sf` object to a `SpatialPoints*` before creating a `ppp` object as shown in the following code chunk. Note that the `maptools` package is required for this step.

```{r}
p.sp  <- as(p.sf, "Spatial")  # Create Spatial* object
p.ppp <- as(p.sp, "ppp")      # Create ppp object
class(p.ppp)
```

If the point layer has an attribute table, its attributes will be converted to `ppp` _marks_.

### Converting a `raster` object to an `im` object (`spatstat`) {-}

The `maptools` package will readily convert a `raster` object to an `im` object using the `as.im.RasterLayer()` function.

```{r}
elev.im <- as.im.RasterLayer(elev.r) # From the maptools package
class(elev.im)
```

## Converting to an `sf` object {-}

All aforementioned spatial formats, except `owin`, can be coerced to an `sf` object via the `st_as_sf` function. for example:

```{r results='hide'}
st_as_sf(p.ppp)  # For converting a ppp object to an sf object
st_as_sf(s.sp)   # For converting a Spatial* object to an sf object
```

## Dissecting the `sf` file object {-}

```{r}
head(s.sf,3)
```

The first line of output gives us the geometry type, `MULTIPOLYGON`,  a multi-polygon data type. This is also referred to as a multipart polygon. A single-part `sf` polygon object will adopt the `POLYGON` geometry.

The next few lines of output give us the layer's bounding extent in the layer's native coordinate system units. You can extract the extent via the `extent()` function as in `extent(s.sf)`.

The next lines indicate the coordinate system used to define the polygon feature locations. It's offered in two formats: **epsg** code (when available) and **PROJ4** string format. You can extract the coordinate system information from an `sf` object via:

```{r}
st_crs(s.sf)
```

The output of this call is an object of class `crs`. Extracting the reference system from a spatial object can prove useful in certain workflows.

What remains of the `sf` summary output is the first few records of the attribute table. You can extract the object's table to a dedicated data frame via:

```{r}
s.df <- data.frame(s.sf)
class(s.df)
head(s.df, 5)
```

The above chunk will also create a geometry column. This column is somewhat unique in that it stores its contents as a **list** of geometry coordinate pairs (polygon vertex coordinate values in this example).

```{r}
str(s.df)
```

You can also opt to remove this column prior to creating the dataframe as follows:

```{r}
s.nogeom.df <- st_set_geometry(s.sf, NULL) 
class(s.nogeom.df)
head(s.nogeom.df, 5)
```

## Exporting to different data file formats {-}

You can export an `sf` object to many different spatial file formats such as a shapefile or a geopackage.

```{r eval=FALSE}
st_write(s.sf, "shapefile_out.shp", driver="ESRI Shapefile")  # create to a shapefile 
st_write(s.sf, " s.gpkg", driver="GPKG")  # Create a geopackage file
```

You can see a list of writable vector formats via a call to `subset(rgdal::ogrDrivers(), write == TRUE)`. Only as subset of the output is shown in the following example. Note that supported file formats will differ from platform to platform.

```{r echo=2}
tail(head(
  subset(rgdal::ogrDrivers(), write == TRUE)
, 16), 7)

```


The value in the `name` column is the driver name used in the `st_write()` function.

To export a raster to a data file, use `writeRaster()` from the `raster` package.

```{r eval=FALSE}
writeRaster(elev.r, "elev_out.tif", format="GTiff" ) # Create a geoTiff file
writeRaster(elev.r, "elev_out.img", format="HFA" )  # Create an Imagine raster file
```

You can see a list of writable raster formats via a call to `subset(rgdal::gdalDrivers(), create == TRUE)`. 

```{r echo=2}
tail(head(
  subset(rgdal::gdalDrivers(), create == TRUE)
, 16), 7)

```

The value in the `name` column is the format parameter name used in the `writeRaster()` function.

